---
title: "Lab4 - HW"
author: "Sara Troshani"
date: "2026-10-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data and Sample

The analysis uses the Household Pulse data. I focus on people ages 25 to 55 who have a non-missing work status.

```{r}
library(tidyverse)
library(AER)

load("~/Desktop/R Studio Data/Household Data/d_HHP2020_241.Rdata")

prime_age_laborforce_data <- d_HHP2020_24 %>%
  filter(
    !is.na(work_kind),
    Age >= 25,
    Age <= 55
  )

summary(prime_age_laborforce_data)
```

This sample includes people ages 25 to 55 who have information about their work status. I use this group because they are more likely to be part of the labor force.

## Main Regression

The main regression looks at how income is related to age, gender, education, race, and Hispanic status.

```{r}
model_1 <- lm(
  income_midpoint ~ Age + Gender + Education + Race + Hispanic,
  data = prime_age_laborforce_data
)

summary(model_1)
confint(model_1)
```

The results show a strong relationship between education and income.

College graduates are predicted to have about $67,093 higher income than the less-than-high-school group, holding the other variables constant.

People with an advanced degree are predicted to have about $88,508 higher income than the less-than-high-school group.

Age also has a positive relationship with income. One more year of age is associated with about $1,239 more in predicted income.

Most of the coefficients are statistically significant. The "some high school" group is not statistically significant.

## Predicted Values

I compare predicted income for a Hispanic Black female with an advanced degree and a Hispanic Black female with a high school education.

```{r}
to_be_predicted1 <- data.frame(
  Age = 25:55,
  Gender = "female",
  Education = "adv degree",
  Race = "Black",
  Hispanic = "Hispanic"
)

to_be_predicted1$yhat <- predict(
  model_1,
  newdata = to_be_predicted1
)

head(to_be_predicted1)

to_be_predicted2 <- data.frame(
  Age = 25:55,
  Gender = "female",
  Education = "high school",
  Race = "Black",
  Hispanic = "Hispanic"
)

to_be_predicted2$yhat <- predict(
  model_1,
  newdata = to_be_predicted2
)

head(to_be_predicted2)
```

The predicted income is higher for the person with an advanced degree at every age.

For example, at age 25, predicted income is about $81,549 for the advanced degree group and about $8,149 for the high school group.

Predicted income also increases with age for both groups.

## Predicted Income Graph

```{r}
comparison <- data.frame(
  Age = 25:55,
  Advanced_Degree = to_be_predicted1$yhat,
  High_School = to_be_predicted2$yhat
)

head(comparison)

ggplot(comparison, aes(x = Age)) +
  geom_line(aes(y = Advanced_Degree, color = "Advanced Degree"), linewidth = 1) +
  geom_line(aes(y = High_School, color = "High School"), linewidth = 1) +
  labs(
    title = "Predicted Income by Age and Education",
    x = "Age",
    y = "Predicted Income",
    color = "Education"
  )
```

The graph shows that predicted income goes up with age for both groups.

The advanced degree group stays much higher than the high school group at every age.

## Robust Standard Errors

```{r}
coeftest(model_1, vcovHC(model_1))
```

I also use robust standard errors to check the results.

The standard errors change a little, but the main results stay the same.

Most of the variables are still statistically significant, and the "some high school" group is still not statistically significant.

## Log Income Model

```{r}
model_log <- lm(
  log(income_midpoint) ~ Age + Gender + Education + Race + Hispanic,
  data = prime_age_laborforce_data
)

summary(model_log)

pred_level <- predict(model_1)
pred_log <- exp(predict(model_log))

mean(pred_level)
mean(pred_log)
```

I also run a model using the log of income instead of income in dollars.

The overall results are similar. Education is still strongly related to income.

The average predicted income from the regular model is about $112,008.

The average predicted income after changing the log predictions back into dollars is about $95,028.

The two numbers are different because one model uses income in dollars and the other uses the log of income.

## Female Only Model

```{r}
female_data <- prime_age_laborforce_data %>%
  filter(Gender == "female")

model_female <- lm(
  income_midpoint ~ Age + Education + Race + Hispanic,
  data = female_data
)

summary(model_female)
```

I also run the regression using only females.

Education still has a strong relationship with income.

For females, college graduates are predicted to have about $71,055 higher income than the less-than-high-school group.

Females with an advanced degree are predicted to have about $93,496 higher income than the less-than-high-school group.

This shows that the relationship between education and income is still strong when I only look at females.

## Joint F Tests

```{r}
drop1(model_1, test = "F")
drop1(model_female, test = "F")
```

The F-tests check whether groups of variables are statistically significant together.

In the main model, age, gender, education, race, and Hispanic status are statistically significant.

Education is also jointly significant in the female-only model.

This means education is important in explaining differences in predicted income.

## Conclusion

Overall, the results show a strong relationship between education and income.

People with higher levels of education tend to have higher predicted income, even after controlling for age, gender, race, and Hispanic status.

The main results stay similar when I use robust standard errors, a log income model, and a female-only sample.

These results show a relationship between education and income, but they do not prove that education directly causes the differences in income.